---
title: IHDS Quick Start Guide
author: ''
date: '2020-08-07'
slug: getting-started-with-ihds-data
categories: []
tags:
  - R
---

I recently used data from the India Human Development Survey (IHDS) for a paper on learning outcomes data in India. IHDS is a panel survey led by Sonalde Desai and others at NCAER/UofMd which collected on a wide range of topics from ~42,000 households across India. The survey includes two rounds of data collection, the first round in 2004/5 and the second round in 2011/12. (According to the website, a third round is in the works.)
For those of you working on India-related research, I would highly recommend checking out the IHDS data. Unlike data from the NSSO, IHDS is free, quick and easy to access, and requires little up-front cleaning. In addition, IHDS covers a much wider range of topics than the NFHS such as income, consumption, agriculture, education, and participation in government programmes.

While the IHDS documentation is excellent, it still takes a bit of time to go through it all. I wrote this blog post to serve as a "quick start" guide to help others (and my future self) quickly get up to speed and working on the data. 


# Accessing the data
Accessing IHDS data is pretty straightforward. You can go to the IHDS webpage [here](https://ihds.umd.edu/), click on "Download Data", and follow the appropriate links or you can just click on the links below. Note that you will have to create a login ID. For more information on which specific files to download, see the next section below.

* Click [here](https://www.icpsr.umich.edu/web/ICPSR/studies/22626) for just the first round of data
* Click [here](https://www.icpsr.umich.edu/web/DSDR/studies/36151) for just the second round of data
* Click [here](https://www.icpsr.umich.edu/web/ICPSR/studies/37382/datadocumentation) for both rounds combined


# Data files

After arriving at the data download page, you will see a bunch of different files to download. First, there are separate files for each **unit of observation**. In addition to the individual-level and household-level datasets, there are also files for medical facility-level data, school-level data, village-level data, and a few others.

If you are downloading the panel data, there are also three different types of datasets corresponding to different ways of combining the two rounds of data. Datasets with  **appended** in the name contain all households surveyed in either round, including households in round 1 that couldn't be found in round 2 and vice versa. Datasets with **long** in the name only include households surveyed in both rounds. Similarly, datasets with **wide** in the name also only include households surveyed in both rounds. The only difference between these two datasets is that long datasets contain two rows for each household (corresponding to each round of the survey) while the wide datasets contain only one row and twice the number of variables. (Note that the variables with the "X" prefix are for the first round.)

You may also choose the format you would like to download the data in (SAS/SPSS/Stata/R/etc).



# Documentation
Your download should include both a data guide and a user guide. (You can also find an online version of the data guide [here](https://www.icpsr.umich.edu/web/pages/DSDR/idhs-II-data-guide.html).) The data guide provides a pretty high-level overview of the data while the user guide provides a bit more detail on the data, including a description of various constructed variables. 

If you are looking for more information about how IHDS was conducted, including the nitty gritty of the sampling strategy, check out [this technical paper.](https://ihds.umd.edu/sites/default/files/publications/papers/technical%20paper%201.pdf) 

In addition, your download should also contain questionnaires and codebooks. (These will be in the sub-folders for each different dataset that you downloaded.)


# Quick tip on finding variables in the datasets
The IHDS datasets have a ton of variables and thus it can be a bit challenging to link dataset variables back to questions and vice versa.

I find that a relatively quick way to find relevant variables in the dataset is to first search for keywords in the questionnaire, jot down the relevant question number, and then search for the question number in the codebook.  Note that the variables in the codebooks are organized by subject rather than ordered by question number so they don't follow the same sequence as the questionnaires.

Also note that codebook section **"ED5: HQ19 11.5 Education: Enrolled now"** refers to variable **ED5** in the dataset which comes from **question 11.5** which is on **page 19** of the household questionnaire.



# Sample R code

I have included some basic R code for importing and analyzing the individual-level round 2 dataset. Note that while I use R, I actually prefer to download the data in Stata format to preserve all the variable labels.

## Set paths
```{r, message=FALSE, results='hide'}
library(tidyverse); library(haven); library(survey)
ihds_ind_dir <- "C:/Users/dougj/Documents/Data/IHDS/IHDS 2012/DS0001"
ind_file <- file.path(ihds_ind_dir, "36151-0001-Data.dta")
```


## Import a single row and create a dataframe of variable labels
With a large dataset like IHDS, I find it useful to first import just one row and create a separate dataframe of variable names and variable labels. This allows me to determine which variables to import later.
```{r, message=FALSE, results='hide'}
ihds_one_row <- read_dta(ind_file, n_max = 1)
labels <- map_chr(ihds_one_row, function(x) as.character(attributes(x)$label)[1])
ihds_labels <- tibble(variable = names(ihds_one_row), label = labels)
```

## Import the data
I import just the variables I need. This saves a lot of time.

```{r, message=FALSE, results='hide'}
ihds <- read_dta(ind_file, col_select = c(STATEID, DISTID, PSUID, HHID, HHSPLITID, PERSONID, IDPSU, WT,  TA8B))

# Create a variables for PSU ID and HH ID which are unique for each PSU and household. This is required for the survey package.
ihds <- ihds %>% mutate(psu_expanded = paste(STATEID, DISTID, PSUID, sep ="-"), hh_expanded = paste(STATEID, DISTID, PSUID, HHID, HHSPLITID, sep ="-"))

# drop the one row with missing values for weights
ihds <- ihds %>% filter(!is.na(WT))

# Create variable for whether the respondent has achieved ASER at level 4. 
# This variable is only non-empty for children aged 8-11, but the functions below
# automatically drop all other respondents.
ihds <- ihds %>% mutate(ASER4 = (TA8B ==4)) %>% mutate(State = as_factor(STATEID))
```

## Use the survey package to generate point estimates and confidence intervals
With survey data you have to use a package like "survey" to generate appropriately weighted estimates and confidence intervals which take into account the sampling design.
``` {r}
ihds_svy <- svydesign(id =~ psu_expanded + hh_expanded, weights =~ WT, data = ihds)
svyby(~ASER4, ~State, ihds_svy, svymean, na.rm=TRUE)

```





